Supplemental Information: Remotely sensing river greenhouse gas exchange velocity using the SWOT satellite

### 2022-02-24 supp_v9.Rmd
prepping to rerun with wider k600

## Test S1: Estimating how hydraulically wide SWOT rivers are
To quantify the prevalence of hydraulically-wide, SWOT-observable rivers, we used the dataset of field-measured river hydraulics outlined in Section 2.1. That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves. Here, we describe how this dataset was filtered down to 171,553 measurements and how we quantified what was 'hydraulically wide'.

@@

## Text S2: Gas exchange model derivations
In this text we provide the full algebra to arrive at the four physically-based gas exchange models we tested in this study. Consult Appendix A for all variable definitions. Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds models, respectively. Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed.

@@

*log-law-of-the-wall dissipation*
$$k_{600}=\alpha(\epsilon_S)^{1/4} \mathbf{(S1)}$$
$$k_{600}=\alpha(\frac{U_*^3}{H})^{1/4} \mathbf{(S2)}$$
$$k_{600}=\alpha(\frac{\sqrt{gR_hS}^3}{H})^{1/4} \mathbf{(S3)}$$
$$k_{600}=\alpha(\frac{(gHS)^{3/2}}{H})^{1/4} \mathbf{(S4)}$$
$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(S5)}$$

*form-drag dissipation*
$$k_{600}=\alpha_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(S7)}$$

@@

*log-law-of-the-wall dissipation*
$$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
$$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$

*form-drag dissipation*
$$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gR_hS)^{3/16}H^{3/8} \mathbf{(S16)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$

## Text S3: BIKER hyperparameterization
In this text we explain in detail how BIKER's prior distributions were set for each river.

Prior distributions are defined by their hyperparameters. For BIKER, prior distriubtions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*. It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementanble. This is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].

$A_0_i$ and $n_i$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 3.2.3 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 6 in the main text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

@@

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$). This was recently shown on many of these rivers for SWOT discharge algorithms [@frassonExploringFactorsControlling2021]. However as noted in the main text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.

@@

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 8 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

@@

In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$, as outlined in the main text. Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations used in the literature to predict river velocity and depth from discharge. Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models.

For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to predict $k_{CO_2}$, which finally is used to predict $FCO_2$. Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019. We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively). Finally, we converted from $k_600$ to $k_{CO_2}$ via @raymondScalingGasTransfer2012.

## Figure S1
![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles. Panel (d) is also presented in the main text as Figure 2.](cache/k600_theory/k600Plot_SI.jpg)

## Figure S2
![Figure S2 Timeseries of the biweekly CO2 data from @beaulieuControlsGasTransfer2012. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Dashed black line denotes atmospheric CO2 at 400 uatm.](cache/FCO2/Beaulieu_timeseries.jpg)

## Figure S3
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_noerr.jpg)

## Figure S4
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 16 SWOT-simulated rivers with the Frasson et al. (2021) error model. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/timeseries_err.jpg)

## Table S1
*Table S1: Studies that gas exchange velocity measurements come from. 'Study' refers to the paper from which these measurements were obtained. Any data wrangling was done by those authors. 'Workers' refers to who actually made the measurements. The Raymond et al. (2012) dataset is itself a meta-analysis. Please see that paper for how those measurements were collected, and see the 'Additional Studies' for the workers who actually collected the measurements.*

@@

## Table S2
*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. As is standard, a carrot accent indicates the predicted value.*

|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|**Validation Scheme**|
|-----------|-------|----------|-----------|--------------|-----------------|
|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|	All rivers and all timesteps|
|Root-mean-square-error|RMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\hat{k_{600, i}}-k_{600, i})^2}$ |0|0 to ∞|All rivers and all timesteps|
|Relative root-mean-square error|RRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{k_{600, i}})^2}$ |0|0 to ∞|By river|
|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|By river|
|Relative bias|rBIAS| $\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})$ |0|-∞ to ∞|By river|

@@

|**Name**|**Depth equation**|**Velocity equation**|**Description**|**Reference**|
|--------|------------------|---------------------|---------------|-------------|
|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|

>>>>

## Test S1: Estimating how hydraulically wide SWOT rivers are
To quantify the prevalence of hydraulically-wide, SWOT-observable rivers, we used the dataset of field-measured river hydraulics outlined in Section 2.1. That dataset has over 500,000 discrete measurements of river width, velocity, area, and discharge that were made by the United States Geological Survey (USGS) to calibrate streamgauge rating curves. Here, we describe how this dataset was filtered down to `r HG_swot$n` rivers.

@@

## Text S2: Gas exchange model derivations
In this text we provide the full algebra to arrive at the four physically-based gas exchange models we tested in this study. Consult Appendix A for all variable definitions. Note that $\alpha$ and $\beta$ are statistical parameters obtained via least squares regression for the small-eddy and Reynolds-extension models, respectively. Subscripts denote variants of the same parameter, depending on which $\epsilon$ model is employed (see Main Text).

@@

$\epsilon = \epsilon_S$

$$k_{600}=\alpha(\epsilon_S)^{1/4} \mathbf{(S1)}$$
$$k_{600}=\alpha(\frac{U_*^3}{H})^{1/4} \mathbf{(S2)}$$
$$k_{600}=\alpha(\frac{\sqrt{gR_hS}^3}{H})^{1/4} \mathbf{(S3)}$$
$$k_{600}=\alpha(\frac{(gHS)^{3/2}}{H})^{1/4} \mathbf{(S4)}$$
$$k_{600}=\alpha(gS)^{3/8}H^{1/8} \mathbf{(S5)}$$

$\epsilon = \epsilon_D$

$$k_{600}=\alpha_1(\epsilon_D)^{1/4} \mathbf{(S6)}$$
$$k_{600}=\alpha_1(gS\bar{U})^{1/4} \mathbf{(S7)}$$

@@

$\epsilon = \epsilon_S$

$$k_{600}=\beta_0(\epsilon_S)^{1/4}A_p^{3/4} \mathbf{(S8)}$$
$$k_{600}=\beta (\frac{U_*^3}{H})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S9)}$$
$$k_{600}=\beta U_*^{9/8}H^{1/8} \mathbf{(S10)}$$
$$k_{600}=\beta (gR_hS)^{9/16}H^{1/8} \mathbf{(S11)}$$
$$k_{600}=\beta (gHS)^{9/16}H^{1/8} \mathbf{(S12)}$$
$$k_{600}=\beta (gS)^{9/16}H^{11/16} \mathbf{(S13)}$$

$\epsilon = \epsilon_D$

$$k_{600}=\beta_0(\epsilon_D)^{1/4}A_p^{3/4} \mathbf{(S14)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}((U_*H)^{1/2})^{3/4} \mathbf{(S15)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gR_hS)^{3/16}H^{3/8} \mathbf{(S16)}$$
$$k_{600}=\beta_1(gS\bar{U})^{1/4}(gHS)^{3/16}H^{3/8} \mathbf{(S17)}$$
$$k_{600}=\beta_1(gS)^{7/16}\bar{U}^{1/4}H^{9/16} \mathbf{(S18)}$$

## Text S3: BIKER hyperparameterization
In this text we explain in detail how BIKER's prior distributions were determined for a given river.

Prior distributions are defined by their hyperparameters. For BIKER, prior distriubtions are formalized as truncated normal distributions of the log-transformed terms such that $log(X) \sim  \mathcal{N}(\mu, \sigma^2)$ for $\lambda < log(X) < \gamma$, using prior hyperparameters mean ($\mu$), standard deviation ($\sigma$), and upper ($\gamma$) and lower bounds ($\lambda$) for any parameter *X*. It is important to again stress, as we do in the main text, that BIKER prior hyperparameters are described using only SWOT data to be completely globally implementable. This use of the data to describe the priors is analogous to the 'empirical Bayes method' [@hoffFirstCourseBayesian2009].

$A_0_i$ and $n_i$ prior hyperparameters were assigned following @brinkerhoffConstrainingRemoteRiver2020. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters are assigned to rivers using only the remotely sensed measurements. For this study, we extracted $A_0_{\lambda}$ and $A_0_{\gamma}$ as the 5th and 95th percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600_t}$ hyperparameters to be defined. $k_{600_{\mu}}$ is set by invoking the hydraulic geometry (HG) relationships developed in section 2.2.1 of the main text using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019. We replaced both depth and velocity terms from our gas exchange model (equation 7 in the Main Text) with these HG models, resulting in equation S19 where Q is the mass-conserved streamflow for the river reach.

@@

Obviously, we have no a priori information about Q. So, we use the Q prior that will be globally available when SWOT launches: a mean annual estimate from a water balance model. This is provided with the simulated SWOT data by both @durandIntercomparisonRemoteSensing2016 and @frassonExploringFactorsControlling2021. Using a temporally-invariant estimate of streamflow is the worst case scenario and BIKER's performance will improve with a more informed prior on streamflow (and therefore $k_{600}$). However as noted in the Main Text, our primary goal with this initial validation is to benchmark BIKER's worst case scenario for performance and so we do that here.

@@

In the scenario that BIKER is run on real SWOT data, $\sigma_k_{600}$ must reflect the full uncertainty implicit in equation 9 in the main text. This means we must also account for uncertainty from the $\beta_1$ parameter. Assuming perfect, no-error measurements are made by the SWOT satellite, the $k_{600}$ relative uncertainty is expressed for some set of hydraulic observations as equation S20. We use the 166 hydraulically-wide measurements in our field-measured $k_{600}$ dataset and run 166 different Monte Carlo simulations (each 10,000 runs) to obtain 166 different $\sigma_k_{600}$ terms. We then take the average of those to be a reasonable estimate of $\sigma_k_{600}$. This ultimately provided a value of 1.1 for the log-transformed $\sigma_k_{600}$ that should be used once SWOT launches.

@@

In brief, gas fluxes ($FCO_2$) are only obtainable at the global scale via predictive equations for $k_{CO_2}$ per equation 2. Here, we obtain $FCO_2$ using the in situ $CO_2$ data described in section 3.1 and $k_{CO_2}$ estimates. The $k_{CO_2}$ estimates are calculated using 1) equation 7, 2) in situ streamflow data, and 3) hydraulic scaling equations from the literature that predict river velocity and depth from streamflow. Specifically, these literature equations include the 'Raymond 2012', 'Raymond 2013', and 'Brinkerhoff 2019' models (Table S3).

For example, the 'Raymond 2012' model uses the hydraulic scaling equations from @raymondScalingGasTransfer2012 to predict velocity and depth, which are in turn used to calculate a $k_{CO_2}$, which finally is used to calculate $FCO_2$. Similarly, the 'Raymond 2013' model uses the equations outlined in @raymondGlobalCarbonDioxide2013 and the 'Brinkerhoff 2019' model uses new equations fit to the hydraulics dataset from @brinkerhoffReconcilingAtaStationAtManyStations2019 following Text S1. We chose to include the 'Brinkerhoff 2019' model as the training dataset is far larger than those used in either of the previous two models (Table S3: 104,624 versus 10,837 versus 1,026, respectively). Finally, we converted from $k_600$ to $k_{CO_2}$ following @raymondScalingGasTransfer2012.

## Figure S1
![Figure S1: Four $k_{600}$~hydraulics models tested in this study, with model assumptions noted in the sub-panel titles. Panel (d) is also presented in the main text as Figure 2 and equation 7.](cache/k600_theory/figS1.jpg)

## Figure S2
![Figure S2 Timeseries of the biweekly $CO_2$ data from @beaulieuControlsGasTransfer2012. Sampling took place 2008-2009 in the Ohio River (upstream of Cincinnati, Ohio, United States). Dashed black line denotes atmospheric $CO_2$ at 400 uatm.](cache/FCO2/figS2.jpg)

## Figure S3
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 47 SWOT-simulated rivers with no additional SWOT errors. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS3.jpg)

## Figure S4
![Figure s3: Observed (orange) versus BIKER-inferred (green) timerseries of daily $k_{600}$ for the 16 SWOT-simulated rivers with the Frasson et al. (2021) error model. Note that the y axis is normalized by maximum observed values to compare across rivers.](cache/validation/figS4.jpg)

## Table S1
*Table S1: Studies that gas exchange velocity measurements come from. 'Study' refers to the paper from which these measurements were obtained. Any data wrangling was done by those authors. 'Workers' refers to who actually made the measurements in the field. The Raymond et al. (2012) dataset is itself a meta-analysis. Please see that paper for how those measurements were collected, and see the 'Additional Studies' for the workers who actually collected the measurements.*

@@

## Table S2
*Table S2: Validation metrics used in this study, where Nt is number of observations and i is the specific observation. σ refers to the variance of the sample and μ refers to the mean of the sample. Carrot accents indicate the predicted value.*

|**Description**|**Acronym**|**Definition**|**Ideal Score**|**Possible Range**|
|---------------|-----------|--------------|---------------|------------------|
|Correlation Coefficient|$r$| $\sqrt{1-(\frac{\sum\limits_{i=1}^{N_t}{(k_{600, i}-\hat{k_{600_i}})^2}}{\sum{(k_{600, i}-\bar{k_{600}})^2}})}$ |1|-1 to 1|
|Normalized root-mean-square error|NRMSE| $\sqrt{\frac{1}{N_t}\sum\limits_{i=1}^{N_t}(\frac{\hat{k_{600, i}}-k_{600, i}}{\bar{k_{600, i}}})^2}$ |0|0 to ∞|
|Kling-Gupta Efficiency|KGE|$1 - \sqrt{ (r(k_{600_i}, \hat{k_{600_i}}) - 1)^2 + (\frac{\mu_{\hat{k_{600_i}}} }{\mu_{k_{600_i}}} - 1)^2 + (\frac{ \sigma _{\hat{k_{600_i}}} / \mu _{\hat{k_{600_i}}} }{ \sigma_{k_{600_i}} / \mu_{k_{600_i}}} - 1)^2 }$|1|-∞ to 1|
|Mean Absolute Error|MAE| $\frac{\sum\limits_{i=1}^{N_t}(|\hat{k_{600_i}}-k_{600_i})|)}{n})$ |0|0 to ∞|

@@

|**Name**|**Depth equation**|**Velocity equation**|**Description**|**Reference**|
|--------|------------------|---------------------|---------------|-------------|
|Brinkerhoff 2019|$D=0.265Q^{0.395}$|$\bar{U}=0.284Q^{0.191}$|104,624 measurements made across the United States at streamgauges|@brinkerhoffReconcilingAtaStationAtManyStations2019|
|Raymond 2012|$D=0.409Q^{0.294}$|$\bar{U}=0.194Q^{0.285}$||1,026 measurements across the United States|@raymondScalingGasTransfer2012|
|Raymond 2013|$D=0.409Q^{0.294},  D=0.449Q^{0.37}$|$\bar{U}=0.194Q^{0.285}, \bar{U}=0.346Q^{0.120}$|Average of the Raymond 2012 equation and one using 9,811 measurements at US streamgauges|@raymondGlobalCarbonDioxide2013|
